---
title: if not ifelse or if_else then what else?
author: David Sjoberg
date: '2019-06-16'
slug: if-not-ifelse-or-if_else-then-what-else
categories: []
tags: [hablar, R, dplyr]
description: 'alternative to ifelse in base R and if_else_ in dplyr'
---

```{r, include = F}
knitr::opts_chunk$set(echo = TRUE,
                      warning=F,
                      error=F,
                      message=F)
```

```{r, include=F}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(hablar, kableExtra, lubridate, tidyverse)
```

It is almost impossible to miss the `ifelse()` and `if_else()` functions in R. It's the goto functions when you want to write an if-statement and apply the logic to every element of that vector. The `hablar` package introduces `if_else_()`, giving you a third way to do a vectorised if-else in R. Let's take an example:

```{r}
library(dplyr)
library(hablar)
library(lubridate)

df <- tibble(name = c("John", "Lisa", "Astrid"),
             birthday = ymd("1976-04-15", "1983-02-03", "1883-10-27"))
```

```{r, echo=F}
knitr::kable(df) %>%
  kable_styling(full_width = F)
```

In the data frame above you can see that the birthday of Astrid is in the year 1883, which we argue is an incorrect birthday. The task is to change it to missing.

### Why use `if_else_` from **hablar**?

 With `if_else_()` from `hablar` we can write it as follows.

```{r, eval=F}
df %>% 
  mutate(birthday = if_else_(year(birthday) < 1900, NA, birthday))
```

```{r, echo = F}
df %>% 
  mutate(birthday = if_else_(year(birthday) < 1900, NA, birthday)) %>% 
  knitr::kable() %>%
  kable_styling(full_width = F)
```

It solved our problem by setting the fauly birthday to `NA` (missing value).

### Why not use `if_else` from **dplyr**?

You definetly could, but it forces you to be more specific than necessary. The above example would raise an error. You need to specify which type of NA you want. In `if_else` you need to write:

```{r, eval=F}
df %>% 
  mutate(birthday = if_else(year(birthday) < 1900, as.Date(NA), birthday))
```

```{r, echo = F}
df %>% 
  mutate(birthday = if_else(year(birthday) < 1900, as.Date(NA), birthday)) %>% 
  knitr::kable() %>%
  kable_styling(full_width = F)
```

It does the same job, but is harder to read.

### Why not use `ifelse()` from **base R**?

You should avoid it since it changes your data type, which makes it unreliable. Additionally, it removes attributes. The above cases with `ifelse()` returns integers which represent the amount of days passed since 1970-01-01. Simply confusing.

```{r, eval=F}
df %>% 
  mutate(birthday = ifelse(year(birthday) < 1900, NA, birthday))
```

```{r, echo = F}
df %>% 
  mutate(birthday = ifelse(year(birthday) < 1900, NA, birthday)) %>% 
  knitr::kable() %>%
  kable_styling(full_width = F)
```

#### Read more

For more examples and information install `hablar` and type `vignette("hablar")`.